New strategies for multi-factoral and temporal behavioral analyses

### 2019-09-07 docs/characterization_manuscript.Rmd
new figs

---
title: Peaks and valleys of prolatin-related gene expression during pigeon parental care stages
author:
  - name: Rayna M Harris
    email: rmharris@ucdavis
    affiliation: UC Davis
    corresponding: "rmharris@ucdavis"
  - name: Suzanne Austin
    affiliation: UC Davis
  - name: Andrew Lang
    affiliation: UNH
  - name: Matthew MacManes
    affiliation: UNH
  - name: Rebecca M Calisi
    affiliation: UC Davis           
address:
  - code: UC Davis
    address: UC Davis, Davis, CA
  - code: UNH
    address: UNH, smalltown, NH
abstract: |
  The goal of this research was to provide the most comprehensive gene expression profile of gene expression activity in the HPG axis of male and female rock doves to provide a deeper undertand of how the reproductive axis response to typical behavioral transitions associated with parental care. Here we used RNA sequencing to measure gene activity at 8 stages of the parental care cyle from nest building, to egg incubation, and through nestling care. Non-parental groups were added as controls, but gene expressions differences were assesed between and across all time points. The three tissue have unique signatures, but within each tissue we find minial sex differences. The pitutary samples displacy the msot plasticity in gene expression, with many changes in gene expression mirroring the typical rise and fall of circulating prolatin that peaks when chicks hatch. Our analysis provides new insight into how suites of genes respond in concert to the demands of offspring care. This data can be used to develop and test hypotheses about the mechanism regulating parental care behavior. 
author_summary: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. 
bibliography: mybibfile.bib
output: rticles::plos_article
csl: plos.csl
---

@@

library(cowplot)
```

@@

```{r, out.width = "360px", echo = F}
knitr::include_graphics("../figures/images/PigeonHPGaxis.png")
```

@@

# Materials and Methods
```{r, out.width = "360px", echo = F}
knitr::include_graphics("../figures/images/samplingtimepoints.png")
```

@@

## Table 1: comparison of total DEGs determined by edgeR and DESeq2

## Fig 1: Total number of gene expression differences between each stage

@@

**Fig x. The magnitude of gene expression changes between each parental transition**.
4-5K genes are differentially expressed beteen control birds and their nest buliding conspecifics in all tissues except the male gonad. 500 - 1000 genes are differentially expressed in the female pituitary from mid-late incubation as well as in the female hypothalamus and pituitary from hatch to nestingly care day 5.

@@

```{r, message=FALSE, warning=FALSE, cache=TRUE, echo = F, fig.width= 5}
ggdraw() + draw_image("../figures/sexes/DoveParentsRNAseq_model2.png") 
```

>>>>

---
title: Peaks and valleys of prolatin-related gene expression during pigeon parental care stages
author:
  - name: Rayna M Harris
    email: rmharris@ucdavis
    affiliation: UC Davis
    corresponding: "rmharris@ucdavis"
  - name: Suzanne Austin
    affiliation: UC Davis
  - name: Andrew Lang
    affiliation: UNH
  - name: Matthew MacManes
    affiliation: UNH
  - name: Rebecca M Calisi
    affiliation: UC Davis           
address:
  - code: UC Davis
    address: UC Davis, Davis, CA
  - code: UNH
    address: UNH, smalltown, NH
abstract: |
  The goal of this research was to provide the most comprehensive gene expression profile of gene expression activity in the HPG axis of male and female rock doves to provide a deeper undertand of how the reproductive axis response to typical behavioral transitions associated with parental care. Here we used RNA sequencing to measure gene activity at 8 stages of the parental care cyle from nest building, to egg incubation, and through nestling care. Non-parental groups were added as controls, but gene expressions differences were assesed between and across all time points. The three tissue have unique signatures, but within each tissue we find minial sex differences. The pitutary samples displacy the msot plasticity in gene expression, with many changes in gene expression mirroring the typical rise and fall of circulating prolatin that peaks when chicks hatch. Our analysis provides new insight into how suites of genes respond in concert to the demands of offspring care. This data can be used to develop and test hypotheses about the mechanism regulating parental care behavior. 
author_summary: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. 
bibliography: mybibfile.bib
output: 
  rticles::plos_article:
    df_print: kable
csl: plos.csl
---

@@

library(cowplot)
library(kableExtra)
```

@@

@@

# Materials and Methods

##  Rock dove housing

Sexually mature male and female rock doves were  housed together in rooftop aviaries on the Barnard Campus. Natural male-female pairings permitted, and small observation windows behind each nest box (64 boxes total) facilitate observations nest building, egg laying, chick hatching, and nesting care every morning from 0900-1100 (correct time?). On specific collection days (Fig 1A, needs to be updated), truck blook, the hypothalamus, the pituitary, and the gonads were collected following common protocols used to sample tissue from birds (Fig 1B). Tissue collection, RNA isolation, cDNA library preparation, sequencing, quality control, and read processing were done as described in Calisi et al. 2018, Lang et al 2019, and Austin et al 2019.

```{r, fig.width= 4, echo = F}
a <- ggdraw() + draw_image("../figures/images/samplingtimepoints.png") 
b <- ggdraw() + draw_image("../figures/images/PigeonHPGaxis.png")

plot_grid(a,b, labels = "AUTO", label_size = 8, ncol = 1)
```

We used EdgeR and DESeq2 to calculate statistical differnces in gene expression between each transition. All subsequent analyses were conducting using variance stabilized data from the DESeq2 model.

@@

Thosands of genes were found to be differentially expressed between the non-parental control group and the building group (Table 1, Fig. 1). Within the time course of parental care, fewer than 1000 genes change there expression over the course of a few days, with the most pronounced difference in toltal number of changes occuring in the female pituitary between incubation days 9 and 17 (Table 1, Fig. 1).

\begin{table}[]
\begin{tabular}{llllllll}
\hline
\textbf{Method} & \textbf{Contrast} & \textbf{Hyp F} & \textbf{Hyp M} & \textbf{Pit F} & \textbf{Pit M} & \textbf{Gon F} & \textbf{Gon M} \\  \hline
EdgeR & Cont - Bldg & 14674 & 117 & 14593 & 144 & 14859 & 4 \\
EdgeR & Bldg - Lay & 14742 & 0 & 14593 & 0 & 14836 & 0 \\
EdgeR & Lay - Inc3 & 0 & 0 & 0 & 0 & 19 & 0 \\
EdgeR & Inc3 - Inc9 & 0 & 1 & 0 & 3 & 2 & 0 \\
EdgeR & Inc9 - Inc 17 & 0 & 2 & 57 & 37 & 9 & 0 \\
EdgeR & Inc17 - Hatch & 0 & 0 & 3 & 1 & 3 & 0 \\
EdgeR & Hatch - Nest5 & 1 & 0 & 3 & 0 & 0 & 0 \\
EdgeR & Nest5 - Nest9 & 0 & 0 & 0 & 0 & 24 & 0 \\  \hline
DESeq2 & Cont - Bldg & 3452 & 4499 & 3950 & 4680 & 4597 & 2135 \\
DESeq2 & Bldg - Lay & 0 & 0 & 101 & 0 & 104 & 1 \\
DESeq2 & Lay - Inc3 & 0 & 0 & 18 & 1 & 440 & 1 \\
DESeq2 & Inc3 - Inc9 & 1 & 0 & 0 & 1 & 8 & 0 \\
DESeq2 & Inc9 - Inc 17 & 1 & 0 & 640 & 186 & 1 & 0 \\
DESeq2 & Inc17 - Hatch & 2 & 0 & 3 & 18 & 1 & 0 \\
DESeq2 & Hatch - Nest5 & 166 & 0 & 340 & 3 & 1 & 0 \\
DESeq2 & Nest5 - Nest9 & 0 & 0 & 1 & 23 & 19 & 0 \\ \hline
\end{tabular}
\caption{Comparison of total DEGs determined by edgeR and DESeq2}
\label{tab:my-table}
\end{table}

@@

**Fig 1. The magnitude of gene expression changes between each parental transition**.
Using DESeq 2, we identified 4-5K genes are differentially expressed beteen control birds and their nest buliding conspecifics in all tissues except the male gonad. 500 - 1000 genes are differentially expressed in the female pituitary from mid-late incubation as well as in the female hypothalamus and pituitary from hatch to nestingly care day 5.

@@

```{r, message=FALSE, warning=FALSE, cache=TRUE, echo = F, fig.width= 5}
a <- ggdraw() + draw_image("../figures/sexes/LDA-4.png") 
b <- ggdraw() + draw_image("../figures/sexes/DoveParentsRNAseq_model2.png")

plot_grid(a,b, labels = "AUTO", label_size = 8, ncol = 1)
```

### 2019-09-07 docs/characterization_manuscript.tex
new figs

\includegraphics[width=360px]{../figures/images/PigeonHPGaxis}

@@

\includegraphics[width=360px]{../figures/images/samplingtimepoints}

@@

\hypertarget{table-1-comparison-of-total-degs-determined-by-edger-and-deseq2}{%
\subsection{Table 1: comparison of total DEGs determined by edgeR and
DESeq2}\label{table-1-comparison-of-total-degs-determined-by-edger-and-deseq2}}

\hypertarget{fig-1-total-number-of-gene-expression-differences-between-each-stage}{%
\subsection{Fig 1: Total number of gene expression differences between
each
stage}\label{fig-1-total-number-of-gene-expression-differences-between-each-stage}}

@@

\textbf{Fig x. The magnitude of gene expression changes between each
parental transition}. 4-5K genes are differentially expressed beteen
control birds and their nest buliding conspecifics in all tissues except
the male gonad. 500 - 1000 genes are differentially expressed in the
female pituitary from mid-late incubation as well as in the female
hypothalamus and pituitary from hatch to nestingly care day 5.

@@

\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-7-1.pdf}

>>>>

@@

\hypertarget{rock-dove-housing}{%
\subsection{Rock dove housing}\label{rock-dove-housing}}

Sexually mature male and female rock doves were housed together in
rooftop aviaries on the Barnard Campus. Natural male-female pairings
permitted, and small observation windows behind each nest box (64 boxes
total) facilitate observations nest building, egg laying, chick
hatching, and nesting care every morning from 0900-1100 (correct time?).
On specific collection days (Fig 1A, needs to be updated), truck blook,
the hypothalamus, the pituitary, and the gonads were collected following
common protocols used to sample tissue from birds (Fig 1B). Tissue
collection, RNA isolation, cDNA library preparation, sequencing, quality
control, and read processing were done as described in Calisi et
al.~2018, Lang et al 2019, and Austin et al 2019.

\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-2-1.pdf}

We used EdgeR and DESeq2 to calculate statistical differnces in gene
expression between each transition. All subsequent analyses were
conducting using variance stabilized data from the DESeq2 model.

@@

Thosands of genes were found to be differentially expressed between the
non-parental control group and the building group (Table 1, Fig. 1).
Within the time course of parental care, fewer than 1000 genes change
there expression over the course of a few days, with the most pronounced
difference in toltal number of changes occuring in the female pituitary
between incubation days 9 and 17 (Table 1, Fig. 1).

\begin{table}[]
\begin{tabular}{llllllll}
\hline
\textbf{Method} & \textbf{Contrast} & \textbf{Hyp F} & \textbf{Hyp M} & \textbf{Pit F} & \textbf{Pit M} & \textbf{Gon F} & \textbf{Gon M} \\  \hline
EdgeR & Cont - Bldg & 14674 & 117 & 14593 & 144 & 14859 & 4 \\
EdgeR & Bldg - Lay & 14742 & 0 & 14593 & 0 & 14836 & 0 \\
EdgeR & Lay - Inc3 & 0 & 0 & 0 & 0 & 19 & 0 \\
EdgeR & Inc3 - Inc9 & 0 & 1 & 0 & 3 & 2 & 0 \\
EdgeR & Inc9 - Inc 17 & 0 & 2 & 57 & 37 & 9 & 0 \\
EdgeR & Inc17 - Hatch & 0 & 0 & 3 & 1 & 3 & 0 \\
EdgeR & Hatch - Nest5 & 1 & 0 & 3 & 0 & 0 & 0 \\
EdgeR & Nest5 - Nest9 & 0 & 0 & 0 & 0 & 24 & 0 \\  \hline
DESeq2 & Cont - Bldg & 3452 & 4499 & 3950 & 4680 & 4597 & 2135 \\
DESeq2 & Bldg - Lay & 0 & 0 & 101 & 0 & 104 & 1 \\
DESeq2 & Lay - Inc3 & 0 & 0 & 18 & 1 & 440 & 1 \\
DESeq2 & Inc3 - Inc9 & 1 & 0 & 0 & 1 & 8 & 0 \\
DESeq2 & Inc9 - Inc 17 & 1 & 0 & 640 & 186 & 1 & 0 \\
DESeq2 & Inc17 - Hatch & 2 & 0 & 3 & 18 & 1 & 0 \\
DESeq2 & Hatch - Nest5 & 166 & 0 & 340 & 3 & 1 & 0 \\
DESeq2 & Nest5 - Nest9 & 0 & 0 & 1 & 23 & 19 & 0 \\ \hline
\end{tabular}
\caption{Comparison of total DEGs determined by edgeR and DESeq2}
\label{tab:my-table}
\end{table}

\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}

\textbf{Fig 1. The magnitude of gene expression changes between each
parental transition}. Using DESeq 2, we identified 4-5K genes are
differentially expressed beteen control birds and their nest buliding
conspecifics in all tissues except the male gonad. 500 - 1000 genes are
differentially expressed in the female pituitary from mid-late
incubation as well as in the female hypothalamus and pituitary from
hatch to nestingly care day 5.

@@

@@
